Appendix 4: species distribution model comparison between inlabru and sdmTMB

### 2021-12-15 doc/appendix-binomial-inla.Rmd
more consise comparison with inla

This model will attempt to replicate the model fit in Appendix 1 as an example of how index standardization can also be achieved with inlabru. 
Once again we will use a data set built into the package: trawl survey data for Pacific Cod in Queen Charlotte Sound.
The density units are kg/km^2^. 
<!-- This can be converted to tonnes per km2 by dividing by 1000, which can sometimes scale better. -->

@@

```{r plot-mesh, fig.asp=0.9, out.width="4in"}
plot(spde$mesh)
```

@@

post_inla <- inla.posterior.sample(result = m_inla, n = 400L)

@@

bru_coefs <- m_bru$summary.random$fac[, c("mean", "sd")]
bru_coefs <- m_bru$summary.fixed[, c("mean", "sd")]
summary(m_bru)
```

@@

We can also fit the model with `sdmTMB`. Note that this is a model with only the fixed year effects and a single spatial field, with presence-absence as the response.

@@

post_sdmTMB <- gather_sims(m, n_sims = 2000L)

@@

Comparing the fixed effects between `sdmTMB` and `inlabru` shows the estimates are correlated, but not perfectly so

@@

```{r}
library(dplyr)
library(ggplot2)
year_factor2011_inla <- unlist(lapply(post_inla, function(s) s$latent["year_factor2017:1", 1]))
year_factor2011_tmb <- filter(post_sdmTMB, .variable == "as.factor.year.2017") %>% pull(.value)

ggplot(bind_rows(data.frame(type = "INLA", post = year_factor2011_inla),
  data.frame(type = "sdmTMB", post = year_factor2011_tmb)), aes(post, colour = type)) +
  geom_density()
```

```{r}
ggplot(post_sdmTMB, aes(.value)) + geom_histogram() + facet_wrap(~.variable, scales = "free")
```

```{r}
x <- inla.smarginal(m_inla$marginals.fixed$year_factor2013)
b <- filter(post_sdmTMB, .variable == "as.factor.year.2013") %>% pull(.value)
g2 <- ggplot(as.data.frame(x), aes(x, y)) + geom_line() +
  geom_histogram(data = data.frame(x = b), mapping = aes(x, after_stat(density)),
    inherit.aes = FALSE, binwidth = 0.1, fill = "blue", alpha = 0.2) +
  ggtitle("Year 2013")

x <- inla.smarginal(m_inla$marginals.fixed$year_factor2011)
b <- filter(post_sdmTMB, .variable == "as.factor.year.2011") %>% pull(.value)
g1 <- ggplot(as.data.frame(x), aes(x, y)) + geom_line() +
  geom_histogram(data = data.frame(x = b), mapping = aes(x, after_stat(density)),
    inherit.aes = FALSE, binwidth = 0.1, fill = "blue", alpha = 0.2) +
  ggtitle("Year 2011")

x <- inla.smarginal(m_inla$marginals.fixed$year_factor2015)
b <- filter(post_sdmTMB, .variable == "as.factor.year.2015") %>% pull(.value)
g2015 <- ggplot(as.data.frame(x), aes(x, y)) + geom_line() +
  geom_histogram(data = data.frame(x = b), mapping = aes(x, after_stat(density)),
    inherit.aes = FALSE, binwidth = 0.1, fill = "blue", alpha = 0.2) +
  ggtitle("Year 2015")

x <- inla.smarginal(m_inla$marginals.fixed$year_factor2017)
b <- filter(post_sdmTMB, .variable == "as.factor.year.2017") %>% pull(.value)
g2017 <- ggplot(as.data.frame(x), aes(x, y)) + geom_line() +
  geom_histogram(data = data.frame(x = b), mapping = aes(x, after_stat(density)),
    inherit.aes = FALSE, binwidth = 0.1, fill = "blue", alpha = 0.2) +
  ggtitle("Year 2017")

x <- inla.smarginal(m_inla$marginals.hyperpar$`Range for i`)
range_post <- filter(post_sdmTMB, .variable == "range") %>% pull(.value)
g3 <- ggplot(as.data.frame(x), aes(x, y)) + geom_line() +
  geom_histogram(data = data.frame(x = range_post), mapping = aes(x, after_stat(density)),
    inherit.aes = FALSE, binwidth = 1.5, fill = "blue", alpha = 0.2) +
  ggtitle("Range")

x <- inla.smarginal(m_inla$marginals.hyperpar$`Stdev for i`)
sigma_E_post <- filter(post_sdmTMB, .variable == "sigma_E") %>% pull(.value)
g4 <- ggplot(as.data.frame(x), aes(x, y)) + geom_line() +
  geom_histogram(data = data.frame(x = sigma_E_post), mapping = aes(x, after_stat(density)),
    inherit.aes = FALSE, binwidth = 0.1, fill = "blue", alpha = 0.2) +
  ggtitle("Random Field SD")

x <- inla.smarginal(m_inla$marginals.hyperpar$`GroupRho for i`)
rho_post <- filter(post_sdmTMB, .variable == "ar1_rho") %>% pull(.value)
g5 <- ggplot(as.data.frame(x), aes(x, y)) + geom_line() +
  geom_histogram(data = data.frame(x = rho_post), mapping = aes(x, after_stat(density)),
    inherit.aes = FALSE, binwidth = 0.02, fill = "blue", alpha = 0.2) +
  ggtitle("AR1 rho")

cowplot::plot_grid(g1, g2, g2015, g2017, g3, g4, g5)

# ------------------

```

```

The other way to set up the fixed coefficients with `inlabru` is to use the `fac` argument. Note the location of the parameters changes from to `summary.random$fac`

@@

These estimates are perfectly correlated with the estimates above,

>>>>

In this example, we fit a presence-absence species distribution model with only the fixed year effects and spatiotemporal random fields using AR1.  
Once again we will use a data set built into the package: trawl survey data for Pacific Cod in Queen Charlotte Sound.
The density units are kg/km^2^. 
<!-- This can be converted to tonnes per km2 by dividing by 1000, which can sometimes scale better. -->

@@

```{r plot-mesh, fig.asp=0.9, out.width="3in"}
plot(spde$mesh)
```

@@

@@

# bru_coefs <- m_bru$summary.random$fac[, c("mean", "sd")]
bru_coefs <- m_bru$summary.fixed[, c("mean", "sd")]
summary(m_bru)
```

Comparing the spatial random effects between `INLA` and `inlabru` shows us that the latter is not estimating well.

```{r}
m_inla$summary.hyperpar[, c("mean", "sd")]
m_bru$summary.hyperpar[, c("mean", "sd")]
```

As a result the fixed effect estimates are also quite different.

```{r}
plot(inla_coefs$mean, bru_coefs$mean)
abline(a = 0, b = 1)
```

@@

We can also fit the model with `sdmTMB`. 
The priors only differ in that... TODO?

@@

@@

Comparing the fixed effects between `sdmTMB` and `inla` shows the estimates are very tightly correlated, but not perfectly so

@@

With sdmTMB, we can extract parameter simulations from the joint precision matrix using `spread_sims()` or `gather_sims()` functions.

```{r sdm-sims}
post_sdmTMB <- gather_sims(m, n_sims = 2000L)

post_sdmTMB2 <- post_sdmTMB %>% mutate(variable = factor(`.variable`, 
                    levels = unique(post_sdmTMB$`.variable`), 
                    labels = c("2009", "2011", "2013", "2015", "2017", "Range", "AR1 rho", "Random Field SD")))
```

We then compare the distribution of these sdmTMB simulations to the marginal posterior densities from INLA. 
From INLA, we can extract the marginal posterior densities using `inla.smarginal()` and name them to match the variable names from the sdmTMB `gather_sims()` function.

```{r inla-densities}
# post_inla <- inla.posterior.sample(result = m_inla, n = 400L)

x <- list()
x[[1]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2009) %>% as_tibble() %>% mutate(variable = "as.factor.year.2009")
x[[2]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2011) %>% as_tibble() %>% mutate(variable = "as.factor.year.2011") 
x[[3]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2013) %>% as_tibble() %>% mutate(variable = "as.factor.year.2013") 
x[[4]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2015) %>% as_tibble() %>% mutate(variable = "as.factor.year.2015") 
x[[5]] <- inla.smarginal(m_inla$marginals.fixed$year_factor2017) %>% as_tibble() %>% mutate(variable = "as.factor.year.2017") 
x[[6]] <- inla.smarginal(m_inla$marginals.hyperpar$`Range for i`) %>% as_tibble() %>% mutate(variable = "range") 
x[[7]] <- inla.smarginal(m_inla$marginals.hyperpar$`Stdev for i`) %>% as_tibble() %>% mutate(variable = "sigma_E") 
x[[8]] <- inla.smarginal(m_inla$marginals.hyperpar$`GroupRho for i`) %>% as_tibble() %>% mutate(variable = "ar1_rho") 
post_inla1 <- do.call(rbind, x)

post_inla2 <- post_inla1 %>% mutate(variable = factor(variable, 
                    levels = unique(post_inla1$variable), 
                    labels = c("2009", "2011", "2013", "2015", "2017", "Range", "Random Field SD", "AR1 rho")))
```

```{r compare-plot, out.width="6in"}
ggplot(post_inla2) + 
  geom_histogram(data=post_sdmTMB2, aes(.value, after_stat(density)), binwidth = 0.1, fill = "blue", alpha = 0.2) + 
  geom_line(data=post_inla2, aes(x, y), inherit.aes = F) + 
  facet_wrap(~variable, scales = "free") +
  ggsidekick::theme_sleek()
```

<!-- The other way to set up the fixed coefficients with `inlabru` is to use the `fac` argument. Note the location of the parameters changes from to `summary.random$fac` -->

@@

<!-- These estimates are perfectly correlated with the estimates above, -->
